Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-02 paper/paper.tex

We assume (as \gaia\ suggests LMA REFERENCE HERE) that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood of the $n$th star is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}
with \\
\[
V_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & (\sigma_{\varpi}\,10^{0.2\,J_n^C})^2
\end{bmatrix}
\]
where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$ band uncertainty, and therefore the dominant uncertainty in the transformed absolute magnitude. So we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty.

@@

We need to correct the photometry used in our method, both for the generation of the prior and the likelihood, for dust. With the emergence of 3D dust maps, it is now possible to estimate dust corrections to stars within the Milky Way. The challenge is getting a proper estimate of the distance to the star before inferring $\varpi_{\true}$, since we need a dust correction prior to doing our inference. Dust corrections are particularly non trivial for stars with poor parallax signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. Although the dust model is probabilistic, as well as our distance inference, to move forward, we break from a fully Bayesian framework, in which we might infer the distance and the dust simultaneously (more about this in the discussion). Instead, we correct for dust using a point estimate from the 3D dust map \citep{green15} sampled at a distance estimated from iteratively inferring and then sampling our posterior pdf over $\varpi_{\true}$. In more detail, to apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior, we infer more precise parallaxes to all the stars in our sample. We then use this more precise parallax posterior to get a measurement of dust in the 3D dust map for each star. And repeat. This approach minimizes severe distances, and their associated severe dust corrections, which would occur if using the likelihoods alone. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median, for example, to do a light dust correction, because over correcting for dust can shoot the likelihood into an unphysical, different space of the \cmd\ and bias the parallax inference. For each distance, we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (LMA REF TABLE SFD). We then regenerate our prior using the lightly dust corrected photometry. The dust values we obtain in our final iteration, and which we apply to our photometry, are visualized in galactic coordinates in Figure \ref{fig:dust}. After applying a dust correction, we do not update the uncertainties, but keep them the same as before the dust correction, which is wrong and under representing our uncertainties but this leads, again, to a lighter deconvolution. \BLedit{In other words, the features of our inferred \cmd\ are conservatively broad (broader than if we over-estimated the errors which would have lead to excessive deconvolution and unrealistically narrow features).} We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.

@@

\subsection{Distances to M67}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67}
\label{fig:m67}
\end{figure}

As proof of concept of our demonstration of concept, we inferred the distances to stars in the open cluster M67. It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away. Previous work (cite CBJ) has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5. With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, the SN reaches 5 for the most certain parallax measurements at about 700 pc, right where this cluster lies. So it is a great test case for comparing parallax inferences with or without a prior. We made a window of 1 square degree on the sky with the cluster M67 lying in the middle. The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data on the 2D sky. With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67. Figure \ref{fig:m67} shows the comparison of the likelihoods and posteriors. The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster. The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not obvious at all. The posterior pdf, shown in the bottom row, are more sharply peaked at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods. Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.

>>>>

We assume that the parallax likelihood is Gaussian, as well as the $(J-K)^c_n$ color likelihood. So our full likelihood of the $n$th star is
\begin{equation}
p(\mathbf{y}_n \given \mathbf{Y}_n, I_n) = \mathcal{N}(\mathbf{y}_n \given \mathbf{Y}_n, V_n)
\end{equation}
with \\
\[
V_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & (\sigma_{\varpi}\,10^{0.2\,J_n^C})^2
\end{bmatrix}
\]
where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$ band uncertainty, and therefore the dominant uncertainty in the transformed absolute magnitude. So we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty.

@@

We need to correct the photometry used in our method, both for the generation of the prior and the likelihood, for dust. With the emergence of 3D dust maps, it is now possible to estimate dust corrections to stars within the Milky Way. The challenge is getting a proper estimate of the distance to the star before inferring $\varpi_{\true}$, since we need a dust correction prior to doing our inference. Dust corrections are particularly non trivial for stars with poor parallax signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. Although the dust model is probabilistic, as well as our distance inference, to move forward, we break from a fully Bayesian framework, in which we might infer the distance and the dust simultaneously (more about this in the discussion). Instead, we correct for dust using a point estimate from the 3D dust map \citep{green15} sampled at a distance estimated from iteratively inferring and then sampling our posterior pdf over $\varpi_{\true}$. In more detail, to apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior, we infer more precise parallaxes to all the stars in our sample. We then use this more precise parallax posterior to get a measurement of dust in the 3D dust map for each star. And repeat. This approach minimizes severe distances, and their associated severe dust corrections, which would occur if using the likelihoods alone. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median, for example, to do a light dust correction, because over correcting for dust can shoot the likelihood into an unphysical, different space of the \cmd\ and bias the parallax inference. For each distance, we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from Table 6 of cite{schlafly11}. We then regenerate our prior using the lightly dust corrected photometry. The dust values we obtain in our final iteration, and which we apply to our photometry, are visualized in galactic coordinates in Figure \ref{fig:dust}. After applying a dust correction, we do not update the uncertainties, but keep them the same as before the dust correction, which is wrong and under representing our uncertainties but this leads, again, to a lighter deconvolution. \BLedit{In other words, the features of our inferred \cmd\ are conservatively broad (broader than if we over-estimated the errors which would have lead to excessive deconvolution and unrealistically narrow features).} We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away.
Previous work \citep{cbj15} has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5.
With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, the SN reaches 5 for the most certain parallax measurements at about 700 pc, right where this cluster lies.
So it is a great test case for comparing parallax inferences with or without a prior.
We made a window of 1 square degree on the sky with the cluster M67 lying in the middle.
The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data on the 2D sky.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
Figure \ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not obvious at all.
The posterior pdf, shown in the bottom row, are more sharply peaked at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods.
Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67}
\label{fig:m67}
\end{figure}
